Dependently Typed Multi-stage Calculus

### 2019-11-30 slides/main.tex
Fix typo in speach draft.

\begin{frame}[fragile]{What is Multi-stage Programming}
    A programming paradigm that enables
    \begin{itemize}
        \item Generation of code at run-time
        \item Evaluation of code at run-time
    \end{itemize}
    Applications
    \begin{itemize}
        \item Speeding up programs using run-time information \\
            (a.k.a. run-time specialization)
        \item DSLs
    \end{itemize}
    \note{
        So, what is multi-stage programming? It is a programming paradigm that
        enables generation of code at run-time and evaluation of code at
        run-time simultaneously. Famous applications of multi-stage programming
        is speeding up programs using run-time information, which is also known
        as run-time specialization and DSLs, such as DSL for linear algebra or
        image processing.
    }
\end{frame}

@@

 # let b = <(~a) + (~a)>   (* Embed code *)
 val b : int code = < ((1 + 2) + (1 + 2)) >
 # run b
 - : int = 6
        \end{Verbatim}
    \end{exampleblock}
    \note{
        This is example code of multi-stage programming in MetaOCaml. In line
        one, I make a code value of 1 + 2 and bind it to variable "a". Angle
        brackets are symbols to make a code values. So, "a" is a value of int
        code type. In line three, I run a code value bounded to "a" using "run"
        primitive. As a result, I get integer 3. We can embed code values into
        another code value.  In line 6, I make a code value by embedding the
        code value banded to "a". Tilde symbol is a symbol to embed a code
        value. Then, I bind a code of 1 + 2 + 1 + 2 to variable "b". When I run
        variable "b", we get integer 6 as a result.
    }
\end{frame}

\begin{frame}[fragile]{Application of Multi-stage Programming: vadd-gen}
    Generate efficient vector addition functions for given lengths.
    \begin{exampleblock}{}
    \begin{Verbatim}[numbers=left]
 val vadd-gen : int -> (vec -> vec -> vec) code
 # let vadd3 = vadd-gen 3
 val vadd3 : (vec -> vec -> vec) code
 (* vadd3 = < fun v w ->                 *)
 (*   let (v, w)=(~v, ~w) in             *)
 (*   [v[0]+w[0];v[1]+w[1];v[2]+w[2]] >  *)
    \end{Verbatim}
    \end{exampleblock}
    \note{
        Speeding up programs is important application of multi-stage
        programming as I said in the two before slide. vadd-gen function in
        this slide generates efficient addition function for given lengths.
        vadd-gen function takes a integer as the vector length and return a
        code value of function which takes two fixed length vectors and return
        the sum of them. For example, when I apply 3 to vadd-gen function and
        bind it to vadd3, vadd3 is a code value which takes two 3 length
        vectors. I show the contents of vadd3 in commented out lines. It is
        important that there is no loop. Loop is unrolled at generation of a
        code value and it helps to make an efficient function.
    }
\end{frame}

\begin{frame}[fragile]{The problem of vadd3 generated by vadd-gen}
    Runtime error occurs for vectors of wrong length.
    \begin{exampleblock}{}
    \begin{Verbatim}[numbers=left]
 val vadd3 : (vec -> vec -> vec) code
 # (run vadd3) [1;2;3] [4;5;6]
 - : vec = [5;7;9]
 # (run vadd3) [1;2] [4;5]
 Exception: Failure
    \end{Verbatim}
    \end{exampleblock}
    \note{
        But there is a problem in vadd3. It is that vadd3 raises runtime errors
        for vectors of wrong length. For example, when I apply correct length
        vectors, [1,2,3] and [4,5,6], I can get the sum of them [5,7,9]. But
        when I apply wrong length vectors, [1,2] and [4,5], I get exception at
        run-time.
    }
\end{frame}

@@

 # (run vadd3) [1;2] [4;5]
 [1;2] has type vec 2
 but was expected of type vec 3
 \end{Verbatim}
    \end{exampleblock}
    \note{
        Our approach to the problem is using dependent types. Dependent types
        are types dependent on values and can prevent illegal value of vadd3.
        Using dependent types, we can express the length of vectors in their
        types. For example, a vector [1;2] can be typed as vec 2 and vadd3 can
        be typed code of function which takes values of vec 3 type and return a
        value of vec 3 type. Now, if I apply wrong length vectors, [1;2] and
        [4;5] to vadd3, I get only type error not run-time error.
    }
\end{frame}

\begin{frame}[fragile]{Our Contribution: \LMD}
    We design a multi-stage calculus \LMD by
    introducing dependent types into \LTP [Hanada\&Igarashi'14].
    \begin{itemize}
        \item Features of multi-stage programming
        \item Features of dependent types
    \end{itemize}
    Technically
    \begin{itemize}
        \item Define well-formed kind, kinding, typing and equivalence rules. 
        \item Prove type soundness.
    \end{itemize}
    \note{
        Our contribution of this research is designing lambda-M-D by
        introducing dependent types into multi-stage calculus lambda-T-P by
        Hanada and Igarashi. lambda-T-P have both of features of multi-stage
        programming and features of dependent types. Technically, we define
        well-formed kind, kinding, typing and equivalence rules and prove type
        soundness.
    }
\end{frame}

@@

\begin{frame}{Contents}
    \tableofcontents[currentsection]
    \note{
        From the next slide, I explain existing multi-stage calculus lambda-T-P which we used as our lambda-M-D.
    }
\end{frame}

\begin{frame}[fragile]{Quick Overview of \LTP [Hanada\&Igarashi'14]}
    \begin{table}
        \begin{tabular}{ c | c | c }
            & \LTP & MetaOCaml \\[2mm]
            \hline
            Quote & \( \textbf{let } a = \TB (1 + 2) \) & \verb| let a = <(1+2)> | \\[2mm]
            Unquote & \( (\TB (\TBL a + \TBL a)) \) & \verb| <(~a)+(~a)> | \\[2mm]
            Code type & \( \TW\ \text{int} \) & \verb| int code |
        \end{tabular}
    \end{table}
    \note{
        This is a quick overview of lambda-TP. In lambda-TP, we can make a code
        value with a black triangle which corresponds to angle brackets in
        MetaOCaml. Left-directed black triangle is used to embed a code values
        which corresponds to tilde in MetaOCaml. Code type is represented with
        white triangle.
    }
\end{frame}

@@

    \note{
        Type of lambda-TP is type-level constants for example int or code type
        or function type. Typing judgements of lambda-TP have shape of M is
        typed tau at stage n under environment gamma.
    }
\end{frame}

@@

    Typing judgements and environments of \LTP include a stage.
    \begin{center}
        \( x:\text{int}@1 \vdash x + 1 : \text{int} {\mathbf{@ 1}} \)
    \end{center}
    \note{
        What is a stage? A stage is a level of nested quote surrounding a term
        . For example, this whole term is on stage 0 because there is no black
        triangle around this term. However, the subterm lambda x dot x + 1 is
        on stage 1 because there is one black triangle around the subterm.
        Typing judgements and environments of lambda-TP include a stage.
    }
\end{frame}

\begin{frame}[fragile]{Typing Rules of \LTP}
    Cannot apply a term to a function in a different stage.
    \begin{center}
        {\footnotesize \infrule[\TApp]{
            \G\V M:\sigma \to \tau @ n \andalso
            \G\V N:\sigma @ n
        }{
            \G\V M\ N : \tau @ n
        }}
    \end{center}
    Increase and decrease stage using \( \TB \) and \( \TBL \).
    \begin{center}
        {\footnotesize \infrule[{\TTB}]{
            \G\V M:\tau @ n+1
        }{
            \G\V\TB M:\TW \tau @ n
        } \hfil
        \infrule[{\TTBL}]{
            \G\V M:\TW \tau @ n
        }{
            \G\V\TBL M:\tau @ n + 1
        }}
    \end{center}
    In general, we cannot use terms of one stage at other stages.
    \note{
        Let's have a look on some typing rules of lambda-TP. The application
        rule is T-App. As Simply Typed Lambda Calculus, types of a function and
        an argument must correspond. In addition to it, there is another
        restriction. It is the function and the argument must live in the same
        stage. Typing rules of quote and unquote is T-black triangle and T-left
        directed black triangle. They say that black triangles increase and
        decrease the stage of terms. In general, we cannot use terms of one
        stage at other stages.
    }
\end{frame}

@@

    \begin{exampleblock}{}
    \begin{center}
        \( \textbf{let}\ \underbrace{\text{mul5}}_{\text{stage }0}\ x = 5 * x\ \textbf{in} \TB\ \underbrace{(\% \text{mul5}\ 4)}_{\text{stage }1} \)
    \end{center}
    \end{exampleblock}
    CSP symbol \( \% \) {\bf{lifts up}} mul5 from stage 0 to stage 1. \\
    (In MetaOCaml, variables are lifted up implicitly.)
    \note{
        But we want to use terms at different stage. Cross Stage Persistence
        (CSP) enables it. CSP allow terms at lower stages to be used at higher
        stages. For example, when we define a mul5 function at stage 0, we
        cannot use it other stages. However, when we insert \% which is a
        symbol of CSP, we can use mul5 at stage 1. CSP symbol \( \% \)
        lifts up mul5 from stage 0 to stage 1.
    }
\end{frame}

@@

\begin{frame}[fragile]{Overview of \LMD}
    Extension of multi-stage calculus \LTP with dependent types
    \begin{itemize}
        \item \LMD terms are the same as \LTP.
        \item Types are extended with dependent function type.
        \item There are equivalence rules for dependent types.
    \end{itemize}
    \note{
        This is overview of lambda-MD. Lambda-MD is an extension of multi-stage
        calculus lambda-TP with dependent types. Its term are the same as
        lambda-TP but types are extended with dependent function type. And
        there are equivalence rules because dependent type system need them.
    }
\end{frame}

\begin{frame}[fragile]{Types and Kinds of \LMD}
    Add dependent function types to types of \LTP.
    \begin{block}{Types}
        \begin{tabbing}
            \hspace{5mm} \( \tau \) \= ::= X \hspace{20mm} \= (type-level constants) \\
            \> \( \mid \TW\ \tau \) \> (code type) \\
            \> \( \bm{\mid \Pi x:\tau.\tau} \) \> (dependent function type) \\
            \> \( \bm{\mid \tau\ M} \) \> (application, e.g. \(\text{vec}\ 3\))
        \end{tabbing}
    \end{block}
    \begin{block}{Kinds}
        \( K ::= * \mid \Pi x:\tau.K \)
    \end{block}
    \note{
        These are types and kinds of lambda-MD. We add dependent function type
        to lambda-TP. So, there is a pi type which means dependent function
        type and application such as vec 3 type. There are kinds also in
        lambda-MD in order to classify types into types of values which is
        represented with star symbol and types dependent on terms which is
        represented with pi x dot K.
    }
\end{frame}

\begin{frame}[fragile]{Kinding rules of \LMD}
    Kiding judgements also include stages.
    \begin{center}
        \( \G \vdash \tau :: K @ n \)
    \end{center}
    Because types may contains terms. (e.g. vec n)
    \begin{center}
        {\footnotesize
        \infrule[{\KApp}]{
            \G\V \sigma:: (\Pi x:\tau.K) @ n \andalso \G\V M:\tau @ n
        }{
            \G\V \sigma\ M::K[x\mapsto M] @ n
        } \\[2mm]
        \infrule[{\KAbs}]{
            \G\V \tau :: * @ n \andalso \G,x:\tau@ n \V \sigma::J@n
        }{
            \G\V(\Pi x:\tau.\sigma) :: (\Pi x:\tau.J) @ n
        }}
    \end{center}
    \note{
        Kiding judgements also include stages. Because types may contains terms
        such as vec n. It is natural that types dependent on a value live in
        the same stage of the value. So, application rule, K-App, says that if
        you want to apply term M to type sigma, they must live in the same
        stage n. K-Abs says when sigma is kinded at stage n and variable x in
        the environment is on stage n also, you can abstract x using Pi.
    }
\end{frame}

\begin{frame}[fragile]{CSP in Types}
    Code type \( \TW \) decreases the stage.
    \begin{center}
        \infrule[{\KTW}]{
            \G\V \tau::*@ n + 1
        }{
            \G\V\TW \tau::*@ n
        }
    \end{center}
    So, we need CSP in types also in order to use \( n \) in \( \TW \).
    \begin{exampleblock}{}
        \begin{center}
            \( \lambda n:\text{int}. \text{zeros}(n): \underbrace{\Pi n:\text{int}. \TW (\overbrace{\text{vec}\ \%n}^{\text{stage }1})}_{\text{stage }0} \)
        \end{center}
    \end{exampleblock}
    \note{
        Code type decreases the stage as you can see in K-white triangle rule.
        So we need CSP in types also. For example, let's consider a function
        which takes integer n and returns a vector of length n filled with
        zero. Because n is a variable at stage 0, you must insert \% symbol
        when you use it in code type. But vec is used without \% symbol. I will
        explain the reason in the next slide.
    }
\end{frame}

@@

\begin{frame}[fragile]{vadd-gen in \LMD}
    Type tells us the length of vector.
    \newcommand{\Vn}{\text{vec}\ \%n}
    \newcommand{\Vt}{\text{vec}\ \%3}
    \begin{exampleblock}{}
    \begin{tabbing}
        \hspace{5mm} \= \( \text{vadd-gen} : \Pi n:\text{int}. \TW (\Vn \to \Vn \to \Vn) \) \\[2mm]
        \> \( \textbf{let}\ \text{vadd3} = \text{vadd-gen}\ 3;; \) \\
        \> \( \text{val}\ \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\[2mm]
    \end{tabbing}
    \end{exampleblock}
    Now, vadd3 takes only vectors of length 3.
    \note{
        Let's return back to vadd-gen function.  vadd-gen is changed as follows
        in \LMD.  The type of vadd-gen says that it takes integer n and returns
        a code value of function which takes two \% n length vectors and
        returns a \% n length vector.  Thanks to dependent types we can see it
        takes length 3 vectors.
    }
\end{frame}

\begin{frame}[fragile]{Technical challenge of vadd3}
    Hard to compose with other functions.
    \newcommand{\Vn}{\text{vec}\ \%n}
    \newcommand{\Vpt}{\text{vec}\ \%3}
    \newcommand{\Vt}{\text{vec}\ 3}
    \begin{exampleblock}{}
    \begin{tabbing}
        \hspace{5mm} \= \( \text{val}\ \text{vadd3} : \TW (\Vpt \to \Vpt \to \Vpt) \) \\
        \> \( \text{val}\ \text{double} : \TW (\Vt \to \Vt) \) \\[2mm]
        \> \( \textbf{let}\ \text{add-double}\ v\ w = \) \\
        \> \hspace{5mm} \( \TB (\underbrace{\TBL\text{double}}_{\Vt \to \Vt}\ \underbrace{(\TBL \text{vadd3}\ \TBL v\ \TBL w)}_{\Vpt});; \) \\
    \end{tabbing}
    \end{exampleblock}
    We cannot compose because of type mismatch.
    \note{
        But there is a technical challenge in vadd3. It is vadd3 is hard to
        compose with other functions. For example, as you have seen, vadd3 has
        the type of code which takes two vec \% 3 values and returns vec \% 3
        values. However, double has type of code which takes vec 3. So we
        cannot compose them because \% symbol raises a type error.
    }
\end{frame}

\begin{frame}[fragile]{Solution}
    \newcommand{\Vt}{\text{vec}\ 3}
    \newcommand{\Vpt}{\text{vec}\ \%3}
    Erase the CSP symbol \( \% \) when \(M \) is closed (\QPercent).
    {\footnotesize{
    \begin{center}
        \infrule[{\QPercent}]{
            M \text{ is a closed term.}
        }{
            \G\vdash \% M \E M
        } \\[2mm]
        \infrule[{\TConv}]{
            \G\V M:\tau@n \andalso
            \G\V \tau\equiv \sigma @n
        }{
            \G\V M:\sigma@n
        }
    \end{center}
    }}
    \begin{exampleblock}{}
        \begin{itemize}
            \item \( \text{vec}\ \%3 \E \text{vec}\ 3 \)
            \item \( \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\
                \hspace{15mm} \( (\E  \TW (\Vpt \to \Vpt \to \Vpt)) \)
        \end{itemize}
    \end{exampleblock}
    \note{
        Our solution for this problem is erasing the CSP symbol \( \% \) when
        \(M \) is closed. \QPercent says that if M is a closed term, \% M is
        equivalent to M. And \TConv says we change the type of term to an
        equivalent type. So, vec \% 3 is equivalent to vec 3 and vadd3 can have
        type of code of function which takes two vec 3 values and returns a vec
        3 value. The type is equivalent to the original type of vadd3.
    }
\end{frame}

@@

\begin{frame}[fragile]{Conclusion}
    We introduced dependent types into multi-stage calculus.
    \begin{itemize}
        \item Avoid illegal use of generated code values with dependent types.
            \begin{itemize}
                \item Full dependent types not restricted ones.
            \end{itemize}
        \item Define equivalence rules.
            \begin{itemize}
                \item Especially for CSP
            \end{itemize}
        \item Prove type soundness.
        \item Future work: algorithmic typing
    \end{itemize}
    \note{
        Here is conclusion. We introduced dependent types into multi-stage
        calculus. It can avoid illegal use of generated code values with
        dependent types. And it dependent types is full dependent types not
        restricted ones. We define equivalence rules for dependent types
        especially for CSP. And we proved type soundness. There is a future
        work of algorithmic typing but we have designed almost all rules of
        algorithmic typing.

>>>>

\begin{frame}[fragile]{What is Multi-stage Programming}
    A programming paradigm that enables
    \begin{itemize}
        \item Generation of code at run-time
        \item Evaluation of code at run-time
    \end{itemize}
    Applications
    \begin{itemize}
        \item Speeding up programs using run-time information \\
            (a.k.a. run-time specialization)
        \item DSLs
    \end{itemize}
    \note{
        So, what is multi-stage programming? It is a programming paradigm that
        enables generation of code at run-time and evaluation of code at
        run-time simultaneously. One of famous applications of multi-stage
        programming is speeding up programs using run-time information, which
        is also known as run-time specialization. A DSL is also a famous
        application such as a DSL for linear algebra or image processing.
    }
\end{frame}

@@

 # let b = <(~a) + (~a)>   (* Embed code *)
 val b : int code = < ((1 + 2) + (1 + 2)) >
 # run b
 - : int = 6
        \end{Verbatim}
    \end{exampleblock}
    \note{
        This is example code of multi-stage programming in MetaOCaml. In line
        one, I make a code value of 1 + 2 and bind it to variable "a". Angle
        brackets are symbols to make a code values. So, "a" is a value of int
        code type. In line three, I run a code value bounded to "a" using "run"
        primitive. As a result, I get integer 3. We can embed code values into
        another code value.  In line 6, I make a code value by embedding the
        code value bounded to "a". Tilde symbol is a symbol to embed a code
        value. Then, I bind a code of 1 + 2 + 1 + 2 to variable "b". When I run
        variable "b", we get integer 6 as a result.
    }
\end{frame}

\begin{frame}[fragile]{Application of Multi-stage Programming: vadd-gen}
    Generate efficient vector addition functions for given lengths.
    \begin{exampleblock}{}
    \begin{Verbatim}[numbers=left]
 val vadd-gen : int -> (vec -> vec -> vec) code
 # let vadd3 = vadd-gen 3
 val vadd3 : (vec -> vec -> vec) code
 (* vadd3 = < fun v w ->                 *)
 (*   let (v, w)=(~v, ~w) in             *)
 (*   [v[0]+w[0];v[1]+w[1];v[2]+w[2]] >  *)
    \end{Verbatim}
    \end{exampleblock}
    \note{
        Speeding up programs is important application of multi-stage
        programming as I said in the two before slide. vadd-gen function in
        this slide generates efficient addition function for given lengths.
        vadd-gen function takes a integer as the length of vectors and return
        code of function which takes two vectors of the length and return the
        sum of them. For example, when I apply 3 to vadd-gen function and bind
        it to vadd3, vadd3 is code which takes two 3 length vectors and returns
        the sum. I show the contents of vadd3 in commented out lines. It is
        important that there is no loop. Loop is unrolled at generation of code
        and it make generated functions efficient.
    }
\end{frame}

\begin{frame}[fragile]{The problem of vadd3 generated by vadd-gen}
    Runtime error occurs for vectors of wrong length.
    \begin{exampleblock}{}
    \begin{Verbatim}[numbers=left]
 val vadd3 : (vec -> vec -> vec) code
 # (run vadd3) [1;2;3] [4;5;6]
 - : vec = [5;7;9]
 # (run vadd3) [1;2] [4;5]
 Exception: Failure
    \end{Verbatim}
    \end{exampleblock}
    \note{
        But there is a problem in vadd3. It is that vadd3 raises runtime errors
        for vectors of wrong lengths. For example, when I apply correct length
        vectors, [1,2,3] and [4,5,6], I can get the sum of them [5,7,9]. But
        when I apply wrong length vectors, [1,2] and [4,5], I get exception at
        run-time.
    }
\end{frame}

@@

 # (run vadd3) [1;2] [4;5]
 [1;2] has type vec 2
 but was expected of type vec 3
 \end{Verbatim}
    \end{exampleblock}
    \note{
        Our approach to the problem is using dependent types. Dependent types
        are types dependent on values and can prevent illegal use of vadd3.
        Using dependent types, we can express the length of vectors in their
        types. For example, a vector [1;2] is as vec 2 and vadd3 is typed code
        of function from vec 3 type to vec 3. Now, if I apply wrong length
        vectors, [1;2] and [4;5] to vadd3, I get only type error not run-time
        error.
    }
\end{frame}

\begin{frame}[fragile]{Our Contribution: \LMD}
    We design a multi-stage calculus \LMD by
    introducing dependent types into \LTP [Hanada\&Igarashi'14].
    \begin{itemize}
        \item Features of multi-stage programming
        \item Features of dependent types
    \end{itemize}
    Technically
    \begin{itemize}
        \item Define well-formed kind, kinding, typing and equivalence rules. 
        \item Prove type soundness.
    \end{itemize}
    \note{
        The contribution of our research is designing lambda-M-D by
        introducing dependent types into multi-stage calculus lambda-T-P by
        Hanada and Igarashi. lambda-MD have both of features of multi-stage
        programming and features of dependent types. Technically, we define
        well-formed kind, kinding, typing and equivalence rules and prove type
        soundness.
    }
\end{frame}

@@

\begin{frame}{Contents}
    \tableofcontents[currentsection]
    \note{
        From the next slide, I explain existing multi-stage calculus lambda-TP
        which is the base of our lambda-MD.
    }
\end{frame}

\begin{frame}[fragile]{Quick Overview of \LTP [Hanada\&Igarashi'14]}
    \begin{table}
        \begin{tabular}{ c | c | c }
            & \LTP & MetaOCaml \\[2mm]
            \hline
            Quote & \( \textbf{let } a = \TB (1 + 2) \) & \verb| let a = <(1+2)> | \\[2mm]
            Unquote & \( (\TB (\TBL a + \TBL a)) \) & \verb| <(~a)+(~a)> | \\[2mm]
            Code type & \( \TW\ \text{int} \) & \verb| int code |
        \end{tabular}
    \end{table}
    \note{
        This is a quick overview of lambda-TP. In lambda-TP, we can make a code
        value with a black triangle, quote, which corresponds to angle brackets
        in MetaOCaml. Left-directed black triangle, unquote, is used to embed a
        code values which corresponds to tilde in MetaOCaml. Code type is
        represented with white triangle.
    }
\end{frame}

@@

    \note{
        Type of lambda-TP is type-level constants for example int or code type
        or function type. Typing judgements of lambda-TP is that M is typed tau
        at stage n under environment gamma.
    }
\end{frame}

@@

    Typing judgements and environments of \LTP include a stage.
    \begin{center}
        \( x:\text{int}@1 \vdash x + 1 : \text{int} {\mathbf{@ 1}} \)
    \end{center}
    \note{
        What is a stage? A stage is a level of nested quote surrounding a term.
        For example, this whole term is on stage 0 because there is no quote
        around this term. However, the subterm lambda x dot x + 1 is on stage 1
        because there is one quote around the subterm.  Typing judgements and
        environments of lambda-TP include a stage.
    }
\end{frame}

\begin{frame}[fragile]{Typing Rules of \LTP}
    Cannot apply a term to a function in a different stage.
    \begin{center}
        {\footnotesize \infrule[\TApp]{
            \G\V M:\sigma \to \tau @ n \andalso
            \G\V N:\sigma @ n
        }{
            \G\V M\ N : \tau @ n
        }}
    \end{center}
    Increase and decrease stage using \( \TB \) and \( \TBL \).
    \begin{center}
        {\footnotesize \infrule[{\TTB}]{
            \G\V M:\tau @ n+1
        }{
            \G\V\TB M:\TW \tau @ n
        } \hfil
        \infrule[{\TTBL}]{
            \G\V M:\TW \tau @ n
        }{
            \G\V\TBL M:\tau @ n + 1
        }}
    \end{center}
    In general, we cannot use terms of one stage at other stages.
    \note{
        Let's have a look on some typing rules of lambda-TP. The application
        rule is T-App. As Simply Typed Lambda Calculus, types of a function and
        an argument must correspond. In addition to it, there is another
        restriction. It is that the function and the argument must live in the
        same stage. Typing rules of quote and unquote is T-quote and T-unquote.
        They say that black triangles increase and decrease the stage of terms.
        In general, we cannot use terms of one stage at other stages.
    }
\end{frame}

@@

    \begin{exampleblock}{}
    \begin{center}
        \( \textbf{let}\ \underbrace{\text{mul5}}_{\text{stage }0}\ x = 5 * x\ \textbf{in} \TB\ \underbrace{(\% \text{mul5}\ 4)}_{\text{stage }1} \)
    \end{center}
    \end{exampleblock}
    CSP symbol \( \% \) {\bf{lifts up}} mul5 from stage 0 to stage 1. \\
    (In MetaOCaml, variables are lifted up implicitly.)
    \note{
        But we want to use terms at different stage, sometime. Cross Stage Persistence
        (CSP) enables it. CSP allow terms at lower stages to be used at higher
        stages. For example, when we define a mul5 function at stage 0, we
        cannot use it other stages. However, when we insert \%, which is a
        symbol of CSP, we can use mul5 at stage 1. CSP symbol \( \% \)
        lifts up mul5 from stage 0 to stage 1.
    }
\end{frame}

@@

\begin{frame}[fragile]{Overview of \LMD}
    Extension of multi-stage calculus \LTP with dependent types
    \begin{itemize}
        \item \LMD terms are the same as \LTP.
        \item Types are extended with dependent function type.
        \item There are equivalence rules for dependent types.
    \end{itemize}
    \note{
        This is overview of lambda-MD. Lambda-MD is an extension of multi-stage
        calculus lambda-TP with dependent types. Its terms are the same with
        lambda-TP but types are extended with dependent function type. And
        there are equivalence rules because dependent type system need them.
    }
\end{frame}

\begin{frame}[fragile]{Types and Kinds of \LMD}
    Add dependent function types to types of \LTP.
    \begin{block}{Types}
        \begin{tabbing}
            \hspace{5mm} \( \tau \) \= ::= X \hspace{20mm} \= (type-level constants) \\
            \> \( \mid \TW\ \tau \) \> (code type) \\
            \> \( \bm{\mid \Pi x:\tau.\tau} \) \> (dependent function type) \\
            \> \( \bm{\mid \tau\ M} \) \> (application, e.g. \(\text{vec}\ 3\))
        \end{tabbing}
    \end{block}
    \begin{block}{Kinds}
        \( K ::= * \mid \Pi x:\tau.K \)
    \end{block}
    \note{
        These are types and kinds of lambda-MD. We add dependent function type
        to lambda-TP. So, there is a pi type which means dependent function
        type and application such as vec 3. There are kinds also in
        lambda-MD in order to classify types into types of values which is
        represented with star symbol and types dependent on terms which is
        represented with pi x dot K.
    }
\end{frame}

\begin{frame}[fragile]{Kinding rules of \LMD}
    Kiding judgements also include stages.
    \begin{center}
        \( \G \vdash \tau :: K @ n \)
    \end{center}
    Because types may contains terms. (e.g. vec n)
    \begin{center}
        {\footnotesize
        \infrule[{\KApp}]{
            \G\V \sigma:: (\Pi x:\tau.K) @ n \andalso \G\V M:\tau @ n
        }{
            \G\V \sigma\ M::K[x\mapsto M] @ n
        } \\[2mm]
        \infrule[{\KAbs}]{
            \G\V \tau :: * @ n \andalso \G,x:\tau@ n \V \sigma::J@n
        }{
            \G\V(\Pi x:\tau.\sigma) :: (\Pi x:\tau.J) @ n
        }}
    \end{center}
    \note{
        Kiding judgements also include stages. Because types may contains terms
        such as n in vec n. It is natural that types dependent on a value live
        in the same stage of the value. So, application rule, K-App, says that
        if you want to apply term M to type sigma, they must live in the same
        stage n. K-Abs says that when sigma is kinded at stage n and variable x
        in the environment is on stage n also, you can abstract x using Pi.
    }
\end{frame}

\begin{frame}[fragile]{CSP in Types}
    Code type \( \TW \) decreases the stage.
    \begin{center}
        \infrule[{\KTW}]{
            \G\V \tau::*@ n + 1
        }{
            \G\V\TW \tau::*@ n
        }
    \end{center}
    So, we need CSP in types also in order to use \( n \) in \( \TW \).
    \begin{exampleblock}{}
        \begin{center}
            \( \lambda n:\text{int}. \text{zeros}(n): \underbrace{\Pi n:\text{int}. \TW (\overbrace{\text{vec}\ \%n}^{\text{stage }1})}_{\text{stage }0} \)
        \end{center}
    \end{exampleblock}
    \note{
        Code type decreases the stage as you can see in K-code.  So we need CSP
        in types also. For example, let's consider a function which takes
        integer n and returns a code of length n vector filled with zeros.
        Because n is a variable at stage 0, you must insert \% symbol when you
        use it in code type. But vec is used without \% symbol. I will explain
        the reason in the next slide.
    }
\end{frame}

@@

\begin{frame}[fragile]{vadd-gen in \LMD}
    Type tells us the length of vector.
    \newcommand{\Vn}{\text{vec}\ \%n}
    \newcommand{\Vt}{\text{vec}\ \%3}
    \begin{exampleblock}{}
    \begin{tabbing}
        \hspace{5mm} \= \( \text{vadd-gen} : \Pi n:\text{int}. \TW (\Vn \to \Vn \to \Vn) \) \\[2mm]
        \> \( \textbf{let}\ \text{vadd3} = \text{vadd-gen}\ 3;; \) \\
        \> \( \text{val}\ \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\[2mm]
    \end{tabbing}
    \end{exampleblock}
    Now, vadd3 takes only vectors of length 3.
    \note{
        Let's return back to vadd-gen function.  vadd-gen is changed as follows
        in \LMD.  The type of vadd-gen says that it takes integer n and returns
        a code value of function which takes two \% n length vectors and
        returns a \% n length vector.  Thanks to dependent types we can see it
        takes only length 3 vectors.
    }
\end{frame}

\begin{frame}[fragile]{Technical challenge of vadd3}
    Hard to compose with other functions.
    \newcommand{\Vn}{\text{vec}\ \%n}
    \newcommand{\Vpt}{\text{vec}\ \%3}
    \newcommand{\Vt}{\text{vec}\ 3}
    \begin{exampleblock}{}
    \begin{tabbing}
        \hspace{5mm} \= \( \text{val}\ \text{vadd3} : \TW (\Vpt \to \Vpt \to \Vpt) \) \\
        \> \( \text{val}\ \text{double} : \TW (\Vt \to \Vt) \) \\[2mm]
        \> \( \textbf{let}\ \text{add-double}\ v\ w = \) \\
        \> \hspace{5mm} \( \TB (\underbrace{\TBL\text{double}}_{\Vt \to \Vt}\ \underbrace{(\TBL \text{vadd3}\ \TBL v\ \TBL w)}_{\Vpt});; \) \\
    \end{tabbing}
    \end{exampleblock}
    We cannot compose because of type mismatch.
    \note{
        But there is a technical challenge in vadd3. vadd3 is hard to
        compose with other functions. For example, as you have seen, vadd3 has
        the type of code which takes two vec \% 3 values and returns vec \% 3
        values. However, double has type of code from vec 3 to vec 3. So we
        cannot compose them because \% symbol raises a type error.
    }
\end{frame}

\begin{frame}[fragile]{Solution}
    \newcommand{\Vt}{\text{vec}\ 3}
    \newcommand{\Vpt}{\text{vec}\ \%3}
    Erase the CSP symbol \( \% \) when \(M \) is closed (\QPercent).
    {\footnotesize{
    \begin{center}
        \infrule[{\QPercent}]{
            M \text{ is a closed term.}
        }{
            \G\vdash \% M \E M
        } \\[2mm]
        \infrule[{\TConv}]{
            \G\V M:\tau@n \andalso
            \G\V \tau\equiv \sigma @n
        }{
            \G\V M:\sigma@n
        }
    \end{center}
    }}
    \begin{exampleblock}{}
        \begin{itemize}
            \item \( \text{vec}\ \%3 \E \text{vec}\ 3 \)
            \item \( \text{vadd3} : \TW (\Vt \to \Vt \to \Vt) \) \\
                \hspace{15mm} \( (\E  \TW (\Vpt \to \Vpt \to \Vpt)) \)
        \end{itemize}
    \end{exampleblock}
    \note{
        Our solution for this problem is erasing the CSP symbol \( \% \) when
        \(M \) is closed. \QPercent says that if M is a closed term, \% M is
        equivalent to M. And \TConv says we change the type of term to an
        equivalent type. So, vec \% 3 is equivalent to vec 3 and vadd3 can be 
        typed of code of function which takes two vec 3 values and returns a vec
        3 value. And the type is equivalent to the original type of vadd3.
    }
\end{frame}

@@

\begin{frame}[fragile]{Conclusion}
    We introduced dependent types into multi-stage calculus.
    \begin{itemize}
        \item Avoid illegal use of generated code values with dependent types.
            \begin{itemize}
                \item Full dependent types not restricted ones.
            \end{itemize}
        \item Define equivalence rules.
            \begin{itemize}
                \item Especially for CSP
            \end{itemize}
        \item Prove type soundness.
        \item Future work: algorithmic typing
    \end{itemize}
    \note{
        Here is conclusion. We introduced dependent types into multi-stage
        calculus. It can avoid illegal use of generated code values with
        dependent types. And its dependent types is full dependent types not
        restricted ones. We define equivalence rules for dependent types
        especially for CSP. And we proved type soundness. There is a future
        work of algorithmic typing but we have designed almost all rules of
        algorithmic typing.
